Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2015-11-25 paint-contour-tree.tex
major structural changes

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%
%%%%%%%%%%Comment out the switch below for full version 
%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\def \STOC {}

\ifx\STOC\undefined
  \newcommand{\FullVer}[1]{#1}
  \newcommand{\SubmitVer}[1]{}
\else
  \newcommand{\FullVer}[1]{}
  \newcommand{\SubmitVer}[1]{#1}
\fi

\documentclass[11pt]{article}
\usepackage[margin=2.54cm]{geometry}
\usepackage[dvipsnames,usenames]{color}
\usepackage{amsfonts,amsmath,amssymb,amsthm,mathtools}
\usepackage{paralist}
\usepackage{algorithmic,algorithm}
\usepackage{bm}
\usepackage{xspace}
\usepackage{centernot}
\usepackage{fancybox}
\usepackage{framed}

%\usepackage{mathabx}
\usepackage[pagebackref,letterpaper=true,colorlinks=true,pdfpagemode=none,urlcolor=blue,linkcolor=blue,citecolor=BrickRed,pdfstartview=FitH]{hyperref}
\usepackage{xspace,prettyref}
\usepackage{color}
\usepackage{graphics}
%\usepackage{MnSymbol}
%% For hyperlinks. Should always be the last package.
%\usepackage[colorlinks,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\let\pref=\prettyref
\newcommand{\mathcalavehyperref}[2]{\texorpdfstring{\hyperref[#1]{#2}}{#2}}
\newcommand{\comment}[1]{ {\color{BrickRed} \footnotesize[#1]}\marginpar{\footnotesize\textbf{\color{red} To Do!}}}
\newcommand{\ignore}[1]{}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemmata}[theorem]{Lemmata}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{subclaim}[theorem]{Subclaim}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{observation}[theorem]{Observation}

@@

%%Ben Added
\DefineNamedColor{named}{RedViolet} {cmyk}{0.07,0.90,0,0.34}
\providecommand{\AlgorithmI}[1]{{\textcolor[named]{RedViolet}{\texttt{\bf{#1}}}}}
\providecommand{\Algorithm}[1]{{\AlgorithmI{#1}\index{algorithm!#1@{\AlgorithmI{#1}}}}}
\newcommand{\zip}{\Algorithm{ZIP}}
\newcommand{\link}{\Algorithm{LINK}}
\newcommand{\sizes}{\Algorithm{SIZES}}
\newcommand{\AMT}{\ensuremath{\mathsf{AMT}}\xspace}
\newcommand{\MT}{\ensuremath{\mathsf{MT}}\xspace}
\newcommand{\BCS}{\ensuremath{\mathsf{BCS}}\xspace}
\newcommand{\sub}{\mathsf{sub}}
\newcommand{\Merge}{\mathsf{Merge}}
\newcommand{\stack}{\mathsf{S}}
\newcommand{\AQ}{\mathsf{AQ}}

@@

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\fcol}{full}
\newcommand{\cur}{cur}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\costs}{\mathop{costs}}
\newcommand{\h}{att}
\newcommand{\init}{{\tt init}}
\newcommand{\jc}{\cJ_C}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pal}{P}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

@@

\newcommand{\XSays}[2]{{
      {$\rule[-0.12cm]{0.2in}{0.5cm}$\fbox{\tt
            #1:} }
      \textcolor{red}{#2}
      \marginpar{\textcolor{blue}{#1}}
      {$\rule[0.1cm]{0.3in}{0.1cm}$\fbox{\tt
            end}$\rule[0.1cm]{0.3in}{0.1cm}$}
      }
   }
\newcommand{\Ben}[1]{{\XSays{Ben}{#1}}}
\newcommand{\remove}[1]{}
\newcommand{\pth}[2][\!]{#1\left({#2}\right)}

@@

\author{
  Benjamin Raichel\BenThanks{}
  \and
  C. Seshadhri\thanks{University of California, Santa Cruz; {\tt scomandu@ucsc.edu}}
}

\title{%A Mountaintop View Requires Minimal Sorting:
Avoiding the Global Sort: 
\break A Faster Contour Tree Algorithm%
\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
}
\date{}

@@

\maketitle
\thispagestyle{empty}

@@

\newpage

\pagenumbering{arabic}

@@

It is convenient to think of $f$ as a manifold sitting in $\RR^{d+1}$,
%  or alternately consider
% $\MM$ in $\RR^{d+1}$
with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. More generally, these are the events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.

@@

Suppose we have $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points.
It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree
is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

\begin{theorem} 
\label{thm:main-alg} 
Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies
that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \to \RR$, where the contour tree $T$
has maximum degree $3$. There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).

@@

\paragraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
characterize hard instances.
% explain \emph{when}
% computing contour trees is expensive, and provide insight into the hardness of particular instances. 
%It is intriguing
%that the cost (as defined earlier) roughly captures the hardness of an instance.
Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
This is tangentially related to the concept of instance-optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
of easy to hard instances. In general, our analysis forces one to look closely
at the relationship between sorting and the contour tree problem.

@@

We detail the basic definitions about contour trees. We follow the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear (PL) function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
except for specially designated \emph{boundary facets}. So $f$ is defined only on the vertices of $\MM$,
and all other values are obtained by linear interpolation.

@@

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, including the infinite facet, which is required to be a boundary facet.  However, 
we allow for other boundary facets, to capture the resulting surface pieces after our algorithm make a horizontal cut.

@@

\begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x| f(x) = h\}$.
A \emph{contour} is a maximal connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
\end{definition}

@@

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but alternate definitions of the contour tree
in~\cite{csa-cctad-00} imply this is a tree. 
Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

\subsection{Some technical remarks}
Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.

As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees, and is the smallest tree that contains all the topological changes
of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.

\section{A tour of the new contour tree algorithm} \label{sec:approach}
Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
can be explained clearly.  
\SubmitVer{Therefore, in the remaining few allotted pages, 
rather than providing an incomplete subset of the technical content, we provide a high level view of the entire result, 
which we view as essential for understanding our approach. The technical details can be found in the full version of the paper \cite{rs-mvrms-14}.}
\FullVer{Therefore, here we first provide a high level view of the entire result, 
which we view as essential for understanding our approach. }

In the interest of presentation, the definitions and theorem statements in this section will
slightly differ from those in the main body\SubmitVer{ of the full version}. They may also differ from the original definitions proposed
in earlier work.

% Our final algorithm is quite technical and has numerous moving parts. This section
% provides a high level view of the entire result, and highlights the main ideas
% and intuition. 
% \SubmitVer{The full version with all the details can be found in the appendix.}
% It is helpful to keep the $d=2$ case in mind, so the input is just a triangulated
% terrain. In the interest of presentation, the definitions and theorem statements in this section will
% slightly differ from those in the main body. They may also differ from the original definitions proposed
% in earlier work.

\paragraph{Do not globally sort:} Arguably, the starting point for this work is \Fig{relative}. We have two terrains
with exactly the same contour tree, but different orderings of (heights of) the critical points.
Turning it around, we cannot deduce the full height ordering of critical points from the contour tree.
Sorting all critical points is computationally unnecessary for constructing the contour tree.
In \Fig{sorting}, the contour tree consists of two balanced binary trees, one of the joins,
another of the splits. Again, it is not necessary to know the relative ordering between the mounds
on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
information is necessary: on the left, the little valleys are higher than the big central valley,
and this is reflected in the contour tree. Leaf paths in the contour tree have points
in sorted order, but incomparable points in the tree are unconstrained.
How do we sort exactly what is required, without knowing the contour tree in advance?

\subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%
%%%%%%%%%%Comment out the switch below for full version 
%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%
\def\SoCGVer{TRUE}%
%%%
%\def\HideComments{TRUE}%
%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\def\ProofInAppendix{TRUE}

\ifx\SoCGVer\undefined
   \newcommand{\InSoCGVer}[1]{}%
   \newcommand{\InNotSoCGVer}[1]{#1}%
\else
%   \def\ProofInAppendix{TRUE}
   \newcommand{\InSoCGVer}[1]{#1}%
   \newcommand{\InNotSoCGVer}[1]{}%
\fi

\newcommand{\CiteFullVer}{\cite{rs-mvrms-14}}%

% \InNotSoCGVer{%
%    \newcommand{\SeeFullVer}{}%
%    \newcommand{\LemmaProofInFullVer}[1]%
%    {%
%       \begin{lemma}
%           #1
%       \end{lemma}%
%    }
% }
% 
% \InSoCGVer{%
%    \newcommand{\SeeFullVer}{{{\CiteFullVer}}}%
%    \newcommand{\LemmaProofInFullVer}[1]%
%    {%
%       \begin{lemma}[For proof see  \CiteFullVer{}]
%           #1
%       \end{lemma}
%    }   
% }

\InSoCGVer{%
   \documentclass[a4paper,UKenglish]{lipics}
%   \documentclass[11pt]{article}%

%  \usepackage{proofapnd}
}
\InNotSoCGVer{%
  \documentclass[11pt]{article}
  \usepackage[margin=2.54cm]{geometry}
}

\InNotSoCGVer{
   \newcommand{\myparagraph}[1]{\paragraph{#1}}
}
\InSoCGVer{
   \newcommand{\myparagraph}[1]{\smallskip\noindent{\textbf{#1}}}
}

\usepackage{amsfonts,amsmath,amssymb,amsthm,mathtools}
\usepackage{paralist}
\usepackage{algorithmic,algorithm}
\usepackage{bm}
\usepackage{xspace}
\usepackage{centernot}
\usepackage{fancybox}
\usepackage{framed}
\usepackage{xspace}
\usepackage{prettyref}
%\usepackage[dvipsnames,usenames]{color}
\usepackage{color}
\usepackage{graphics}
%\usepackage[pagebackref,colorlinks=true,urlcolor=blue,linkcolor=blue,citecolor=BrickRed,pdfstartview=FitH]{hyperref}
\usepackage{hyperref}%
\hypersetup{%
   breaklinks,%
   colorlinks=true,%
   urlcolor=[rgb]{0.2,0.0,0.0},
   linkcolor=[rgb]{0.5,0.0,0.0},%
   citecolor=[rgb]{0,0,0.445},
   filecolor=[rgb]{0,0,0.4},
   anchorcolor=[rgb]{0.0,0.1,0.2}
}

%\usepackage{MnSymbol}
%% For hyperlinks. Should always be the last package.
%\usepackage[colorlinks,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\let\pref=\prettyref
\newcommand{\mathcalavehyperref}[2]{\texorpdfstring{\hyperref[#1]{#2}}{#2}}
\newcommand{\comment}[1]{ {\color{BrickRed} \footnotesize[#1]}\marginpar{\footnotesize\textbf{\color{red} To Do!}}}
\newcommand{\ignore}[1]{}

\InNotSoCGVer{
 \newtheorem{theorem}{Theorem}[section]
}
\newtheorem{lemmata}[theorem]{Lemmata}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{subclaim}[theorem]{Subclaim}
\newtheorem{proposition}[theorem]{Proposition}
\InNotSoCGVer{
 \newtheorem{lemma}[theorem]{Lemma}
 \newtheorem{corollary}[theorem]{Corollary}
 \newtheorem{example}[theorem]{Example}
 \newtheorem{definition}[theorem]{Definition}
}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{question}[theorem]{Question}
\theoremstyle{definition}
\InNotSoCGVer{ 
 \newtheorem{remark}[theorem]{Remark} 
}
\newtheorem{observation}[theorem]{Observation}

@@

%%Ben Added
\DefineNamedColor{named}{RedViolet} {cmyk}{0.07,0.90,0,0.34}
\providecommand{\AlgorithmI}[1]{{\textcolor[named]{RedViolet}{\texttt{\bf{#1}}}}}
\providecommand{\Algorithm}[1]{{\AlgorithmI{#1}\index{algorithm!#1@{\AlgorithmI{#1}}}}}
\newcommand{\zip}{\Algorithm{ZIP}}
\newcommand{\link}{\Algorithm{LINK}}
\newcommand{\sizes}{\Algorithm{SIZES}}
\newcommand{\AMT}{\ensuremath{\mathsf{AMT}}\xspace}
\newcommand{\MT}{\ensuremath{\mathsf{MT}}\xspace}
\newcommand{\BCS}{\ensuremath{\mathsf{BCS}}\xspace}
\newcommand{\sub}{\mathsf{sub}}
\newcommand{\Merge}{\mathsf{Merge}}
%\newcommand{\stack}{\mathsf{S}}
\newcommand{\AQ}{\mathsf{AQ}}

@@

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\fcol}{full}
\newcommand{\cur}{cur}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\costs}{\mathop{costs}}
\newcommand{\h}{att}
\newcommand{\init}{{\tt init}}
\newcommand{\jc}{\cJ_C}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pal}{P}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\stack}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

\newcommand{\pathTree}{P_{\mathcal{S}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}

@@

\newcommand{\XSays}[2]{{
      {$\rule[-0.12cm]{0.2in}{0.5cm}$\fbox{\tt
            #1:} }
      \textcolor{red}{#2}
      \marginpar{\textcolor{blue}{#1}}
      {$\rule[0.1cm]{0.3in}{0.1cm}$\fbox{\tt
            end}$\rule[0.1cm]{0.3in}{0.1cm}$}
      }
   }
\ifx\HideComments\undefined%
\else%
\renewcommand{\XSays}[2]{}%
\fi%

\newcommand{\Ben}[1]{{\XSays{Ben}{#1}}}
\newcommand{\remove}[1]{}
\newcommand{\pth}[2][\!]{#1\left({#2}\right)}

@@

@@

\InNotSoCGVer{
\author{
  Benjamin Raichel%\BenThanks{}
  \and
  C. Seshadhri%\thanks{University of California, Santa Cruz; {\tt scomandu@ucsc.edu}}
}

\title{%A Mountaintop View Requires Minimal Sorting:
Avoiding the Global Sort: 
\break A Faster Contour Tree Algorithm%
\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
}
\date{}
}

\InSoCGVer{
% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{%A Mountaintop View Requires Minimal Sorting:
Avoiding the Global Sort: 
\break A Faster Contour Tree Algorithm%
\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
}
\titlerunning{Avoiding the Global Sort: A Faster Contour Tree Algorithm} %optional, in case that the title is too long; the running title should fit into the top page column

\author[1]{Benjamin Raichel}
\author[2]{C. Seshadhri}
  \affil[1]{%
      Department of Computer Science, %
      University of Texas at Dallas\\ %
      Richardson, TX, 75080, USA\\ %
      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}} %
   }%
   \affil[2]{%
      Department of Computer Science, %
      University of California, Santa Cruz\\ %
      Santa Cruz, CA, 95064, USA\\ %
      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
   }%

\authorrunning{B. Raichel and C. Seshadhri} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

\Copyright{Benjamin Raichel and Seshadhri Comandur}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\subjclass{F.2.2, I.1.2, I.3.5}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
\keywords{contour trees, computational topology, computational geometry}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \serieslogo{}%please provide filename (without suffix)
% \volumeinfo%(easychair interface)
%   {Billy Editor and Bill Editors}% editors
%   {2}% number of editors: 1, 2, ....
%   {Conference title on which this volume is based on}% event
%   {1}% volume
%   {1}% issue
%   {1}% starting page number
% \EventShortName{}
% \DOI{10.4230/LIPIcs.xxx.yyy.p}% to be completed by the volume editor
% % Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
}

\maketitle
\InNotSoCGVer{
\thispagestyle{empty}
}

@@

\InNotSoCGVer{
\newpage
\pagenumbering{arabic}
}

@@

It is convenient to think of $f$ as a manifold sitting in $\RR^{d+1}$,
%  or alternately consider
% $\MM$ in $\RR^{d+1}$
with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. More generally, these are the events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.\footnote{%
Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
For $d>2$, some subsequent works, such as \cite{csa-cctad-00}, only include changes in the number of components.
}
\Ben{See added footnote}

@@

Suppose we have $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
\Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

\begin{theorem} 
\label{thm:main-alg} 
%Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies that all critical points have distinct values. 
%Our algorithm does not need this, but it simplifies the proof and presentation.} 
\Ben{Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the contour tree $T$
has maximum degree $3$}.\footnote{The degree $\leq 3$ assumption helps simplify the proof and presentation, and is often made, as in \cite{kobps-ctsssit-97}.}\Ben{See footnote} 
There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).

@@

\myparagraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
characterize hard instances.
% explain \emph{when}
% computing contour trees is expensive, and provide insight into the hardness of particular instances. 
%It is intriguing
%that the cost (as defined earlier) roughly captures the hardness of an instance.
Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
This is tangentially related to the concept of instance-optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
of easy to hard instances. In general, our analysis forces one to look closely
at the relationship between sorting and the contour tree problem.

@@

We detail the basic definitions about contour trees. We follow the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
except for specially designated \emph{boundary facets}. So $f$ is defined only on the vertices of $\MM$,
and all other values are obtained by linear interpolation.

@@

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, including the infinite facet, which is required to be a boundary facet.  However, 
we allow for other boundary facets, to capture the resulting surface pieces after our algorithm makes a horizontal cut.

@@

\begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x| f(x) = h\}$.
A \emph{contour} is a connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
\end{definition}

@@

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but alternate definitions of the contour tree
in~\cite{csa-cctad-00} imply this is a tree. 
Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
%
\Ben{Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and the total degree is at most $3$.}
%
%We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
%This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\InSoCGVer{See Appendix~\ref{sec:techmarks} for further technical remarks on the above definitions.}
\newcommand{\technicalRemarks}{
%\subsection{Some technical remarks}
Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.

As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees, and is the smallest tree that contains all the topological changes
of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.
}
\InNotSoCGVer{
\subsection{Some technical remarks}
\technicalRemarks
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\intuitionSection}{
\section{A tour of the new contour tree algorithm} \label{sec:approach}
\InNotSoCGVer{
Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
can be explained clearly.  Therefore, here we first provide a high level view of the entire result.
}
\InSoCGVer{
For the $d=2$ case, where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
can be explained clearly.  Therefore, here we provide a high level view of the entire result.
}

In the interest of presentation, the definitions and theorem statements in this section will
slightly differ from those in the main body. They may also differ from the original definitions proposed
in earlier work.
